---
title: "Viz II - Visualization with ggplot2"
date: September 30, 2025
output: github_document
---

```{r }
library(tidyverse)
library(patchwork)
library(p8105.datasets)
```

```{r}
data("weather_df")
  
```

Let's make a basic scatterplot. We can use xlabs and ylabs to label the axis.Aadd labels if you're sending to someone else

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.5) +
  labs(
      x = "Minimun daily temp",
      y = "Maximun daily temp",
      title = "Temperature scatterplot",
      caption = "Data from NOAA",
      color = "Location"
  )
```

We are able to control the scales in the graphs. Scale x continous because oru x-axis is contrinuous but can also be categotical. We can labels and breaks. We can do a square root transformaiton on the y axis. We can also add limits to zoom in the plots

We can also do filter to place limits rather than using the limits function. This can be done before typing ggplot. For example, filter(tmin > 10, <tmax 30 ).

We can alos playing around with the color pallete using scale_color_hue, but don't do that. Refer to color palletes used from online sources. Look at the p8105 website for reference. Use the viridis package instead using the viridis::scale_color_viridi

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.5) +
  labs(
      x = "Minimun daily temp",
      y = "Maximun daily temp",
      title = "Temperature scatterplot",
      caption = "Data from NOAA",
      color = "Location"
  ) +
  scale_x_continuous(
    breaks = c(-20, 0, 25),
    labels = c("-20C", "0", "25")
  ) +
  scale_y_continuous(
    trans = "sqrt",
    limits = c(10, 30)
  ) +
  viridis::scale_color_viridis(
    discrete = TRUE
  )
  
```

Themes help with changing everything in the plot. 

We can also use ggsave to save the plot or create a variable to save the plot in the workspace/environment for manipulation letter

```{r}
gg_temperature = weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.5) +
  labs(
      x = "Minimun daily temp",
      y = "Maximun daily temp",
      title = "Temperature scatterplot",
      caption = "Data from NOAA",
      color = "Location"
  ) +
  scale_x_continuous(
    breaks = c(-20, 0, 25),
    labels = c("-20C", "0", "25")
  ) +
  scale_y_continuous(
    trans = "sqrt",
    limits = c(10, 30)
  ) +
  viridis::scale_color_viridis(
    discrete = TRUE
  )
```

Themes help with changing everything in the plot. You can use theme_minimal, theme_bw, theme_dark to change the theme of the plot. legend.posotion moves the legends to the bottom of the X axis. These_minimal we get white backgrounds with gray lines

```{r}
gg_temperature + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

Let's add data in geoms. 

```{r}
central_park_df =
  weather_df |> 
  filter(name == "CentralPark_NY")

molokai_df =
  weather_df |> 
  filter(name == "Molokai_HI")

ggplot(data = molokai_df, aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line(data = central_park_df)
```

Let's make three plots and combine using patchwork

```{r}
ggp_tmax_tmin =
  weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color= name)) +
  geom_point(alpha = 0.5) +
  theme(legend.position = "bottom")

ggp_prec_density =
  weather_df |> 
  filter(prcp > 0) |> 
  ggplot(aes(x = prcp, fill = name)) +
  geom_density(alpha = 0.5) +
  theme(legend.position = "none")

ggp_temp_season =
  weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom")
```

Let's combine plots side by side using + and / to change positions in the same place to convey as much information as possible

```{r}
(ggp_tmax_tmin + ggp_prec_density) / ggp_temp_season
```

Let's now do some data manipulation. Factor variables > rads a string like centrla park but it reads a number. Factor are in alphabetical order. But maybe you want things in a particular order. We need to convince ggplot that we want certain things as a factor. We need to move things arund using the mutate function. Take a look below. Factor relevel asings molokai 1, centralpark 2, and waterhole 3 

```{r}
weather_df |> 
  mutate(name = fct_relevel(name, c("Molokai_HI", "CentralPark_NY", "Waterhole_WA"))) |> 
  ggplot(aes(x = name, y = tmax, fill= name)) +
  geom_violin(alpha = 0.5)
```


Factor order tells R which variable you're operating on and which things you're reodering it in respect to

```{r}
weather_df |> 
  mutate(name = fct_reorder(name, tmax)) |> 
  ggplot(aes(x = name, y = tmax, fill= name)) +
  geom_violin(alpha = 0.5)
```

Factor in order, basically means what order the data appears in the dataset

```{r}
pulse_df =
  haven::read_sas("./data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  ) |> 
  mutate(visit = fct_inorder(visit))

pulse_df |> 
  ggplot(aes(x = visit, y = bdi)) +
  geom_boxplot()
```

